My title

### 2024-09-24 paper/paper.qmd
write Discussion in paper and add some reference

library(tidyverse)
library(ggplot2)
library(dplyr)
library(scales) 
library(tidyr)

@@

# Data {#sec-data}
## Overview {#sec-data-overview}
Some of our data is of penguins (@fig-bills), from @palmerpenguins.

@@

@@

cleaned_shelter_data <- read.csv("/Users/liz/Downloads/starter_folder-main/data/analysis_data/shelter_analysis_data.csv")

# Ensure the date column is in the correct format (Month-Year format)
cleaned_shelter_data$Date <- as.Date(paste0("01-", cleaned_shelter_data$Date), format="%d-%b-%y")

@@

# Load ggplot2 for plotting
ggplot(total_population_data, aes(x = Date, y = Homeless_Count)) +
  geom_line(color = "blue", linewidth = 1.5) +  # Adjust line color and width
  labs(x = "Date",
       y = "Homeless Population Count") +
  scale_x_date(date_labels = "%Y", date_breaks = "1 year") +  
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

@@

@fig-total-population illustrates the fluctuation in Toronto's homeless population from 2018 to 2024. The chart shows cyclical rises and falls, with more pronounced declines around 2020 and 2021, followed by a steady recovery. The population reaches its highest point in 2024. Throughout this period, the homeless population varies between approximately 7,500 and 11,000 individuals.

@@

# Summarize data for all years (total homeless population by age group)
total_data <- cleaned_shelter_data %>%
  summarise(Age_Under_16 = sum(Age_Under_16, na.rm = TRUE),
            Age_16_24 = sum(Age_16_24, na.rm = TRUE),
            Age_25_34 = sum(Age_25_34, na.rm = TRUE),
            Age_35_44 = sum(Age_35_44, na.rm = TRUE),
            Age_45_54 = sum(Age_45_54, na.rm = TRUE),
            Age_55_64 = sum(Age_55_64, na.rm = TRUE),
            Age_65_Over = sum(Age_65_Over, na.rm = TRUE))

# Reshape data for plotting
total_data_long <- tidyr::pivot_longer(total_data, 
                                       cols = starts_with("Age"), 
                                       names_to = "Age_Group", 
                                       values_to = "Count")

# Calculate percentages
total_count <- sum(total_data_long$Count)
total_data_long <- total_data_long %>%
  mutate(Percent = round(Count / total_count * 100))

# Create a pie chart with percentage labels
ggplot(total_data_long, aes(x = "", y = Count, fill = Age_Group)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar("y") +
  geom_text(aes(label = paste0(Percent, "%")), 
            position = position_stack(vjust = 0.5), size = 4) +
  theme_void() +
  theme(legend.position = "right") +
  scale_fill_brewer(palette = "Set3")  # Use a color palette for clarity

@@

```{r}
#| echo: false
#| warning: false
#| label: fig-gender-distribution
#| fig-cap: "Homeless Population by Gender in Toronto (2018-2024)"
#| 
# Read the cleaned data
cleaned_shelter_data <- read.csv("/Users/liz/Downloads/starter_folder-main/data/analysis_data/shelter_analysis_data.csv")

# Ensure the date column is in the correct format
cleaned_shelter_data$date <- as.Date(paste0("01-", cleaned_shelter_data$Date), format="%d-%b-%y")
cleaned_shelter_data$year <- as.numeric(format(cleaned_shelter_data$date, "%Y"))  # Extract year as numeric

# Summarize data by year for different genders
gender_data <- cleaned_shelter_data %>%
  group_by(year) %>%
  summarise(Male = sum(Male_Count, na.rm = TRUE),
            Female = sum(Female_Count, na.rm = TRUE),
            Non_Binary = sum(Transgender_Non_Binary_Count, na.rm = TRUE))

# Calculate total population for each year and for overall gender proportions
total_population <- gender_data %>%
  summarise(Male = sum(Male), Female = sum(Female), Non_Binary = sum(Non_Binary))

# Calculate percentages for each gender over the entire period
total_population_long <- pivot_longer(total_population, 
                                      cols = c("Non_Binary", "Female", "Male"), 
                                      names_to = "Gender", 
                                      values_to = "Total_Count")

# Calculate proportions
total_population_long <- total_population_long %>%
  mutate(Percent = round(Total_Count / sum(Total_Count) * 100, 1))

# Create a string for the percentages to use in the legend
gender_percentages <- paste0(total_population_long$Gender, " (", total_population_long$Percent, "%)")

# Reshape data for plotting
gender_data_long <- pivot_longer(gender_data, 
                                 cols = c("Male", "Female", "Non_Binary"), 
                                 names_to = "Gender", 
                                 values_to = "Count")

gender_data_long$Gender <- factor(gender_data_long$Gender, levels = c("Non_Binary", "Female", "Male"))

# Plot area chart with percentages in the legend
ggplot(gender_data_long, aes(x = year, y = Count, fill = Gender)) +
  geom_area(position = "stack") +
  labs(x = "Year", y = "Homeless Population Count", fill = "Gender") +
  scale_fill_manual(values = c("Male" = "blue", "Female" = "pink", "Non_Binary" = "purple"),
                    labels = gender_percentages) +  # Add percentages to the legend
  theme_minimal()

@@

@fig-gender-distribution presents the gender distribution of Toronto's homeless population from 2018 to 2024. During this period, males have consistently made up the largest proportion, accounting for approximately 61.8% of the total homeless population. Females make up about 36.8%, while non-binary individuals make up about 1.4%. Over time, the total homeless population of each gender has shown similar trends, meaning that the proportion of each gender has remained almost constant.

@@

# Summarize data by year for Chronic and All Population
chronic_data <- cleaned_shelter_data %>%
  filter(Population_Group == 'Chronic') %>%
  group_by(year) %>%
  summarize(homeless_count = sum(Homeless_Count))

@@

# Plot the stacked bar chart with Chronic at the top and add percentages
ggplot(combined_data, aes(x = year, y = homeless_count, fill = group)) +
  geom_bar(stat = "identity", position = "stack") +
  geom_text(data = proportions, aes(x = year, y = homeless_count_chronic, label = paste0(percentage, "%")),
            vjust = 1.5, color = "white", size = 4) +  # Add percentage labels in white color
  labs(x = "Year", y = "Homeless Population Count", fill = "Group") +
  theme_minimal() +
  scale_fill_manual(values = c("Chronic" = "blue", "Non-Chronic" = "orange"))

@@

The @chronic-population displays the changes in the number and proportion of chronic and non-chronic homeless people in Toronto between 2018 and 2024. The number of chronically homeless has been growing over this time, with over 30,000 people in 2018, and by 2023, the number has gone to about 60,000, nearly doubling. the slight drop in numbers in 2024 is due to the fact that the statistics were last updated on September 17, 2024, and it is projected that by the end of the year, the number will have exceeded the 2023 one. Similarly, the percentage of chronically homeless is growing every year, from 29.7% in 2019 to 63.7% in 2024. Note that in 2019 to 2020 and 2023 to 2024, the rate grows faster, by about 10 percent.

@@

## First discussion point {#sec-first-point}

If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.

## Second discussion point

## Third discussion point

@@

# Model details {#sec-model-details}

>>>>

library(tidyverse)
library(ggplot2)
library(dplyr)
library(tidyr)

@@

# Data {#sec-data}
## Overview {#sec-data-overview}
The dataset used for this analysis is the “Toronto Shelter System Flows” dataset from the Toronto Open Data Platform [@torontosheltersystem]. It shows the total number of homeless people entering and exiting shelters each month from 2018 to September 2024, as well as the number of different age groups and genders. It also shows the number and proportion of chronic homeless, refugees, youth, single adults, and families in the total homeless population. These data are updated monthly and reflect the latest trends in Toronto's homeless population and shelter system[@torontosheltersystem].

The variables of interest in this analysis vary by age group (Under 16, 16-24, 25-34, 35-44, 45-54, 55-64, 65 and over) and gender (Male, Female, Transgender/Non-Binary), as well as Chronic and All populations. According to @torontosheltersystem, Chronic refers to those who, according to the federal government's definition of chronic homelessness, meet one of the following two criteria: first, stayed in a shelter for at least 180 nights in the past year; and second, stayed in a shelter multiple times in the past three years, totaling at least 546 cumulative nights. All populations refers to the total number of homeless persons covered in the report without any demographic breakdown.

Similar datasets available on the City of Toronto's Open Data Catalogue include Daily Shelter & Overnight Service Occupancy & Capacity and Daily Shelter Occupancy[@opendata]. However, these datasets only provide an overview of shelter usage without categorizing the homeless population using the shelters. As a result, they were not used in this analysis.

@@

```{r}
#| echo: false
#| warning: false
#| message: false

# Loading the Cleaned Dataset #
cleaned_shelter_data <- read.csv("/Users/liz/Downloads/starter_folder-main/data/analysis_data/shelter_analysis_data.csv")

# Generated by ChatGPT
cleaned_shelter_data$Date <- as.Date(paste0("01-", cleaned_shelter_data$Date), format="%d-%b-%y")
cleaned_shelter_data$year <- as.numeric(format(cleaned_shelter_data$Date, "%Y"))

@@

@@

# This code is based on knowledge from Alexander (2023).
# Load ggplot2 for plotting
ggplot(total_population_data, aes(x = Date, y = Homeless_Count)) +
  geom_line(color = "blue", linewidth = 1.5) +  
  labs(x = "Date",
       y = "Homeless Population Count") +
  scale_x_date(date_labels = "%Y", date_breaks = "1 year") +  
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

@@

@fig-total-population illustrates the fluctuation in Toronto's homeless population from 2018 to 2024. The chart shows cyclical rises and falls, with more pronounced declines around 2020 and 2021, followed by a steady increase.The population reaches its highest point in 2024. Throughout this period, the homeless population varies between approximately 7,500 and 11,000 individuals.

@@

# Summarize data for all years (total homeless population by age group)
# Generated by ChatGPT
total_data <- cleaned_shelter_data %>%
  summarise(Age_Under_16 = sum(Age_Under_16, na.rm = TRUE),
            Age_16_24 = sum(Age_16_24, na.rm = TRUE),
            Age_25_34 = sum(Age_25_34, na.rm = TRUE),
            Age_35_44 = sum(Age_35_44, na.rm = TRUE),
            Age_45_54 = sum(Age_45_54, na.rm = TRUE),
            Age_55_64 = sum(Age_55_64, na.rm = TRUE),
            Age_65_Over = sum(Age_65_Over, na.rm = TRUE))

# Reshape data for plotting
# Generated by ChatGPT
total_data_long <- tidyr::pivot_longer(total_data, 
                                       cols = starts_with("Age"), 
                                       names_to = "Age_Group", 
                                       values_to = "Count")

# Calculate percentages
# Generated by ChatGPT
total_count <- sum(total_data_long$Count)
total_data_long <- total_data_long %>%
  mutate(Percent = round(Count / total_count * 100))

# Create a pie chart with percentage labels
# This code is based on knowledge from Alexander (2023) and further refined using the RStudio Help documentation.
ggplot(total_data_long, aes(x = "", y = Count, fill = Age_Group)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar("y") +
  geom_text(aes(label = paste0(Percent, "%")), 
            position = position_stack(vjust = 0.5), size = 4) +
  theme_void() +
  theme(legend.position = "right") +
  scale_fill_brewer(palette = "Set3")

@@

```{r}
#| echo: false
#| warning: false
#| label: fig-gender-distribution
#| fig-cap: "Homeless Population by Gender in Toronto (2018-2024)"
#| 
# Summarize data by year for different genders
# This logic is similar to the previous graph of age group 
gender_data <- cleaned_shelter_data %>%
  group_by(year) %>%
  summarise(Male = sum(Male_Count, na.rm = TRUE),
            Female = sum(Female_Count, na.rm = TRUE),
            Other = sum(Transgender_Non_Binary_Count, na.rm = TRUE))

# Calculate total population for each gender over the entire period
total_population <- gender_data %>%
  summarise(Male = sum(Male), Female = sum(Female), Other = sum(Other))

# Reshape data for plotting
total_population_long <- pivot_longer(total_population, 
                                      cols = c("Other", "Female", "Male"), 
                                      names_to = "Gender", 
                                      values_to = "Total_Count")

# Calculate percentages for each gender
total_population_long <- total_population_long %>%
  mutate(Percent = round(Total_Count / sum(Total_Count) * 100, 1))

# Create a pie chart similar to the one used for age groups
# This code is based on knowledge from Alexander (2023) and further refined using the RStudio Help documentation.
ggplot(total_population_long, aes(x = "", y = Total_Count, fill = Gender)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar("y") +
  geom_text(aes(label = paste0(Percent, "%")), 
            position = position_stack(vjust = 0.5), size = 4, fontface = "bold", color = "gray30") +  
  theme_void() +
  theme(legend.position = "right") +
  scale_fill_manual(values = c("Male" = "#7FB3D5", "Female" = "#F5B7B1", "Other" = "#D2B4DE"))

@@

@fig-gender-distribution presents the gender distribution of Toronto's homeless population from 2018 to 2024. During this period, males have consistently made up the largest proportion, accounting for approximately 61.8% of the total homeless population. Females make up about 36.8%, while non-binary individuals make up about 1.4%.

@@

# Summarize data by year for Chronic and All Population
# Generated by ChatGPT
chronic_data <- cleaned_shelter_data %>%
  filter(Population_Group == 'Chronic') %>%
  group_by(year) %>%
  summarize(homeless_count = sum(Homeless_Count))

@@

# Plot the stacked bar chart with Chronic at the top and add percentages
ggplot(combined_data, aes(x = year, y = homeless_count, fill = group)) +
  geom_bar(stat = "identity", position = "stack") +
  geom_text(data = proportions, aes(x = year, y = homeless_count_chronic, label = paste0(percentage, "%")),
            vjust = 1.5, color = "white", size = 3) +  # Add percentage labels in white color
  labs(x = "Year", y = "Homeless Population Count", fill = "Group") +
  theme_minimal() +
  scale_fill_manual(values = c("Chronic" = "blue", "Non-Chronic" = "orange"))

@@

The @fig-chronic-population displays the changes in the number and proportion of chronic and non-chronic homeless people in Toronto between 2018 and 2024. The number of chronically homeless has been growing over this time, with over 30,000 people in 2018, and by 2023, the number has gone to about 60,000, nearly doubling. the slight drop in numbers in 2024 is due to the fact that the statistics were last updated on September 17, 2024, and it is projected that by the end of the year, the number will have exceeded the 2023 one. Similarly, the percentage of chronically homeless is growing every year, from 29.7% in 2019 to 63.7% in 2024. Note that in 2019 to 2020 and 2023 to 2024, the rate grows faster, by about 10 percent.

@@

## Overview of Homeless Trends (2018-2024)
In @sec-data-results，trends in Toronto's homeless population from 2018 to 2024 are presented, focusing on key factors such as age, gender, and chronic versus non-chronic conditions. The homeless population fluctuates significantly, being significantly impacted in specific years such as 2020 and 2021[@fig-total-population], with males and the 35-44 age group making up the majority of the homeless population[@fig-age-distribution,@fig-gender-distribution], and chronic homelessness showing an increasing trend[@fig-chronic-population].
## Impact of COVID-19 on Homelessness
The homeless population declined significantly in 2020 and 2021, likely due to COVID-19. To meet social distancing requirements, the Toronto government reduced shelter capacity and opened temporary shelters, including additional spaces in hotels and motels[@may2023scan]. These actions may explain the drop during these years. However, the rebound in homelessness afterward suggests that issues like housing affordability and lack of social support remain unresolved, highlighting the need for sustainable long-term solutions.
## Demographic and Chronic Homelessness Insights
An analysis of age, gender, and chronic homelessness provides key insights into Toronto's homeless population. Homelessness is concentrated among working-age adults, especially those aged 35-44 and 25-34, likely due to challenges with employment or housing stability. In terms of gender, males consistently make up the majority, while non-binary individuals account for only 1.4%. This suggests shelters should consider different designs for men and women, while also addressing the needs of non-binary individuals, despite their smaller numbers. The rise in chronic homelessness suggests that existing measures are not enough to help people exit homelessness permanently.

@@
